(Deep) Induction for GADTs\vspace*{-0.1in}

### 2021-04-15 induction-paper/ENTCS-GADT-induction.tex
work on Section 4, added  Enrico's work on later sections

  \begin{adjustbox}{varwidth=6.8in, max width=6.6in, margin=-0.1in 0in
      -0.2in 0in, fbox, center} 
{\small
\[\begin{array}{lll}
\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
  (var\;s\,T_A) \, liftA} & = & \mathsf{cvar \, A\, Q_A\, s\, T_A\,
  liftA}\\ 
\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
  (abs \,B \,C \,e \,s \,T_B \, t_C) \, (Q_B , Q_C , liftE,
  lift_{T_B}, lift_{t_C})} & = & \mathsf{cabs\,A\,B\,C\, Q_A\,
  Q_B\, Q_C\, e\, s\, T_B\, t_C\, liftE\, lift_{T_B}\, p_C}\\
\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
    (app \,B \,\,t_{BA} \, t_B)\, (Q_B , list_{t_{BA}}, list_{t_B})} &
= & \mathsf{capp\,A\,B\,Q_A\, Q_B\, t_{BA}\, t_B\, p_{BA} \, p_B}\\
  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\,
    (list \,B \,e \, ts) \, (Q_B , liftE', lift_{List})} & = & 
  = \mathsf{clistc \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE'\, p_{List} }
\end{array}\]}

@@

{\color{red} HERE!!!!}

%Since, as seen in Section~\ref{sec:GADT-nested},
%GADTs featuring nesting do not admit an induction rule,
%we restrict our scope to GADTs not featuring nesting.
Finally, we generalize the approach taken in the previous examples and
provide a general framework to derive deep induction rules for
arbitrary GADTs.  For that, we need to give a grammar for the types we
will be considering.  A generic GADT
\begin{equation*}
\mathsf{data\ G : Set^\alpha \to Set\ where}\\
\mathsf{\;\;\;\;\;\;\;\;c_i\, :\, F_i\,G\,\ol{B} \to G (\ol{K_i\,\ol{B}})}
\end{equation*}
is defined by a finite number of constructors $\mathsf{c_i}$.
In the definition above, $\mathsf{F_i}$ is a type constructor with signature $\mathsf{(Set^{\alpha} \to Set) \to Set^{\beta} \to Set}$
and each $\mathsf{K_i}$ is a type constructor with signature $\mathsf{Set^{\beta} \to Set}$
(i.e. a type constructor of arity $\mathsf{\beta}$).
The overline notation denotes a finite list:
%$\mathsf{\ol{a}}$ is a list of types of length $\mathsf{\alpha}$,
%so that it can be applied to the type constructor $\mathsf{G}$ of arity $\mathsf{\alpha}$.
%and the vertical bars notation denotes the length of a list:
%$\mathsf{G}$ depends on a finite list of types $\mathsf{\ol{a}}$ of length $\mathsf{\alpha}$,
each of the $\mathsf{\alpha}$-many $\mathsf{K_i}$ is a type constructor of arity $\mathsf{\beta}$
so that it can be applied to the list of types $\mathsf{\ol{B}}$ of length $\mathsf{\beta}$.
%depends on a finite list of types $\mathsf{\ol{b}}$ of length $\mathsf{\beta}$,
Moreover, notice that the arity of $\mathsf{G}$ matches the number of type constructors $\mathsf{K_i}$.
We allow each $\mathsf{F_i}$ to be inductively built in the following ways (and with the following restrictions):
\item $\mathsf{F_i = F_i' \times F_i''}$ where $\mathsf{F_i'}$ and $\mathsf{F_i''}$ have the same signature as $\mathsf{F_i}$ and are built recursively from the same induction rules.
\item $\mathsf{F_i = F_i' + F_i''}$ where $\mathsf{F_i'}$ and $\mathsf{F_i''}$ have the same signature as $\mathsf{F_i}$ and are built recursively from the same induction rules.
\item $\mathsf{F_i = F_i' \to F_i''}$
where $\mathsf{F_i'}$ does not contain the recursive variable,
i.e., $\mathsf{F_i' : Set^\beta \to Set}$ is a type constructor of arity $\mathsf{\beta}$,
and $\mathsf{F_i''}$ has the same signature as $\mathsf{F_i}$ and is built recursively from the same induction rules.
\item $\mathsf{F_i\,G\,\ol{B} = G(\ol{F_j\,\ol{B}})}$
where none of the $\mathsf{\alpha}$-many $\mathsf{F_j}$ contains the recursive variable,
i.e., $\mathsf{F_j : Set^\beta \to Set}$
is a type constructor of arity $\mathsf{\beta}$
for each $\mathsf{j = 1, \dots, \alpha}$.
Such restriction is necessary to prevent nesting,
as that would break the induction rule as discussed in Section~\ref{sec:GADT-nested}.
\item $\mathsf{F_i\,G\,\ol{B} = H\,\ol{B}}$
where $\mathsf{H}$ is a type constructor of arity $\mathsf{\beta}$ not containing the recursive variable,
Notice that this covers the case in which $\mathsf{F_i}$ is a closed type,
so, in particular, the unit and empty types, $\mathsf{1}$ and $\mathsf{0}$,
and the case in which $\mathsf{F_i}$ is a variable.
%The only requirement on $\mathsf{H}$ is that it needs to have a predicate-lifting function
%(and any type made of constant types, sums, products, arrow types and other GADTs satisfies this condition).
\item $\mathsf{F_i\,G\,\ol{B} = H (\ol{F_k\,G\,\ol{B}})}$
and $\mathsf{F_k}$ has the same signature as $\mathsf{F_i}$  and is built recursively from the same induction rules, for every $\mathsf{k = 1 \dots \gamma}$.
Moreover, we require that $\mathsf{H}$ is not a GADT itself (but we allow it to be an ADT or even a nested type).
This way we know that $\mathsf{H}$ admits functorial semantics~\cite{jp20}, and thus there is a map function for $\mathsf{H^{\wedge}}$,
\mathsf{HLMap : \forall (\ol{C : Set}) (\ol{Q_C\;Q_C' : C \to Set})
	\to \ol{PredMap\,C\,Q_C\,Q_C'}
%where $\mathsf{PredMap : \forall (C : Set) \to (C \to Set) \to (C \to Set) \to Set}$
%is defined as
%\mathsf{PredMap\,C\,Q_c\,Q_c' = \forall (x : c) \to Q_c\,x \to Q_c'\,x}
%and represents the type of morphisms between predicates.
A concrete way to define $\mathsf{HLMap}$ is to proceed by induction on the structure of the type $\mathsf{H}$,
and give an inductive definition when $\mathsf{H}$ is an ADT or a nested type.
Such details are not essential to the present discussion, and thus we omit them.
We can summarize the above inductive definition with the following grammar
(but beware that the above restrictions and requirements still apply):
\mathsf{F_i\,G\,\ol{B} :=
F_i'\,G\,\ol{B} \times F_i''\,G\,\ol{B}
\ \vert\ F_i'\,G\,\ol{B} + F_i''\,G\,\ol{B}
\ \vert\ F_i'\,\ol{B} \to F_i''\,G\,\ol{B}
\ \vert\ G(\ol{F_j\,\ol{B}})
\ \vert\ H\,\ol{B}
\ \vert\ H (\ol{F_k\,G\,\ol{B}})
A further requirement that applies to all of the types appearing above,
including the types $\mathsf{K_i}$,
is that every type needs to have a predicate-lifting function.
This is not an overly restrictive condition, though:
all types made by sums, products, arrow types and type application do, and so do GADTs as defined above.
A concrete way to define the predicate-lifting function for a type is to proceed by induction on the structure of the type,
and we have seen in the previous sections examples of how to do so for products and type application.
We do not give here the general definition of lifting,
as that would require to first present a full type calculus,
and that is beyond the scope of the paper.

%Before proceeding with the derivation of the induction rule for a generic GADT,
%we need to define the type
%\[
%\mathsf{PredMap\,a\,Q_a\,Q_a' = \forall (x : a) \to Q_a\,x \to Q_a'\,x}
%\]
%where $\mathsf{Q_a}$ and  $\mathsf{Q_a'}$ are predicates on  $\mathsf{a : Set}$,
%which represents the type of morphisms between predicates $\mathsf{Q_a}$ and  $\mathsf{Q_a'}$.
%Moreover, if $\mathsf{H}$ is a type constructor having an associated predicate-lifting function $\mathsf{H^{\wedge}}$,
%we define a map function $\mathsf{HLMap}$ for the lifting $\mathsf{H^{\wedge}}$ as
%\[
%\mathsf{HLMap : \forall (a : Set) (Q_a\;Q_a' : a \to Set) \to PredMap\,a\,Q_a\,Q_a' \to PredMap\,(G\,a)\,(G^{\wedge}\,a\,Q_a)\,(G^{\wedge}\,a\,Q_a')}
%\]

Consider a generic GADT as defined above,
\begin{equation}\label{eq:gadt}
\begin{array}{l}
\mathsf{data\ G : Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;\;\;c\, :\, F\,G\,\ol{B} \to G (K\,\ol{B})}
\end{array}
\end{equation}
which, for ease of notation,
we assume to be a unary type constructor (i.e. it depends on a single type parameter $\mathsf{A}$)
and to have only one constructor $\mathsf{c}$.
Extending the argument to GADTs of arbitrary arity and with multiple constructors presents no difficulty
other than heavier notation.
In the definition above, $\mathsf{F}$ has signature $\mathsf{(Set \to Set) \to Set^{\beta} \to Set}$
and each $\mathsf{K}$ has signature $\mathsf{Set^{\beta} \to Set}$.
The constructor $\mathsf{c}$ can be rewritten using the $\mathsf{Equal}$ type as
\[
\mathsf{c : \forall (\ol{B : Set}) \to Equal\,A\,(K\,\ol{B}) \to F\,G\,\ol{B} \to G\,A}
\]
which is the form we shall use from now on.

@@

In the previous sections,
we derive induction rules for examples of GADTs that do not feature nesting,
in the sense that their constructors contain no nested calls of the recursive variable,
as truly nested types (such as $\mathsf{Bush}$, Equation~\ref{eq:bush}) do.
Since both nested types
and GADTs without nesting admit induction rules,
as seen in the previous sections,
it is just natural to expect that GADTs with nesting would as well.
Surprisingly, that is not the case:
indeed, the induction rule generally relies on (unary) parametricity of the semantic interpretation,
and in the case of nested types it also relies on functorial semantics~\cite{jp20},
but GADTs cannot admit both functorial and parametric semantics at the same time~\cite{jgj21}.
In this section we show how induction for GADTs featuring nesting goes wrong
by analyzing the following concrete example of such a type.
\begin{equation}\label{gadt-nested}
\begin{array}{l}
\mathsf{data\, G\,(a : Set) : Set\,where}\\
\mathsf{\;\;\;\;\;\;C :\, G(G\,a) \to G(a \times a)}
\end{array}
\end{equation}
The constructor $\mathsf{C}$ can be rewritten as
\[
\mathsf{C : \exists (b : Set) \to Equal\,a\,(b \times b) \to G(G\,b) \to G\,a}
\]
which is the form we shall use from now on.
The predicate-lifting function of $\mathsf{G}$,
\[
\mathsf{G^{\wedge} : \forall (a : Set) \to (a \to Set) \to G\,\ol{a} \to Set}
\]
is defined as
\[
\mathsf{G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)
= \exists (Q_b : b \to Set)
\to Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e
\times G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}
\]
where $\mathsf{Q_a : a \to Set}$, $\mathsf{e : Equal\,a\,(b \times b)}$ and $\mathsf{x : G(G\,b)}$.
Finally, let $\mathsf{CC}$ be the function
\begin{multline*}
\mathsf{\lambda (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set)} \\
\mathsf{\to \forall (a\;b : Set) (Q_a : a \to Set) (Q_b : b \to Set) (e : Equal\,a\,(b \times b)) (x : G\,(G\,b))} \\
\mathsf{\to Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e
	\to P\,(G\,b)\,(P\,b\,Q_b)\,x
	\to P\,a\,Q_a\,(C\,b\,e\,x)}
\end{multline*}
associated with the $\mathsf{C}$ constructor.

The induction rule for $\mathsf{G}$ is
\begin{multline*}
\mathsf{\forall (P : \forall (a : Set) \to (a \to Set) \to G\,a \to Set) \to CC\,P} \\
\mathsf{\to \forall (a : Set)(Q_a : a \to Set)(y : G\,a) \to G^{\wedge}\,a\,Q_a\,y \to P\,a\,Q_a\,y}
\end{multline*}
Consistently with the previous examples,
to validate the induction rule
we try to define a term of the above type, $\mathsf{DIG}$, as
\[
\mathsf{DIG\,P\,cc\,a\,Q_a\,(C\,b\,e\,x)\,(Q_b, L_E, L_G)
= cc\,a\,b\,Q_a\,Q_b\,e\,x\,L_E\,p}
\]
where $\mathsf{cc : CC\,P}$ and
\begin{itemize}
\item $\mathsf{C\,b\,e\,x : G\,a}$,
i.e., $\mathsf{e : Equal\,a\,(b \times b)}$ and $\mathsf{x : G(G\,b)}$;
\item $\mathsf{(Q_b, L_E, L_G) : G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$,
i.e., $\mathsf{Q_b : b \to Set}$, $\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e}$,
and $\mathsf{L_G : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}$.
\end{itemize}
We still need to define $\mathsf{p : P\,(G\,b)\,(P\,b\,Q_b)\,x}$.
We do so by using the induction rule and letting
\[
\mathsf{p = DIG\,P\,cc\,(G\,b)\,(P\,b\,Q_b)\,x\,q}
\]
where we still need to provide $\mathsf{q : G^{\wedge}\,(G\,b)\,(P\,b\,Q_b)\,x}$.
If we had the map function of $\mathsf{G^{\wedge}}$,
\[
\mathsf{GLMap : \forall (a : Set) (Q_a\;Q_a' : a \to Set) \to PredMap\,a\,Q_a\,Q_a' \to PredMap\,(G\,a)\,(G^{\wedge}\,a\,Q_a)\,(G^{\wedge}\,a\,Q_a')}
\]
then we would be able to define
\[
\mathsf{q = GLMap\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,(P\,b\,Q_b)\,(DIG\,P\,cc\,b\,Q_b)\,x\,L_G}
\]
Unfortunately, we cannot define such a $\mathsf{GLMap}$.
Indeed, its definition would have to be
\[
\mathsf{GLMap\,a\,Q_a\,Q_a'\,M\,(C\,b\,e\,x)\,(Q_b, L_E, L_G) = (Q_b', L_E', L_G')}
\]
where $\mathsf{Q_a : a \to Set}$, $\mathsf{Q_a' : a \to Set}$,
$\mathsf{M : PredMap\,a\,Q_a\,Q_a'}$,
$\mathsf{C\,b\,e\,x : G\,a}$, i.e.,
\begin{itemize}
\item $\mathsf{e : Equal\,a\,(b \times b)}$;
\item $\mathsf{x : G\,(G\,b)}$;
\end{itemize}
$\mathsf{(Q_b, L_E, L_G) :G^{\wedge}\,a\,Q_a\,(C\,b\,e\,x)}$, i.e.,
\begin{itemize}
\item $\mathsf{Q_b : b \to Set}$;
\item $\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e}$;
\item $\mathsf{L_G : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x}$;
\end{itemize}
and $\mathsf{(Q_b', L_E', L_G') : G^{\wedge}\,a\,Q_a'\,(C\,b\,e\,x)}$, i.e.,
\begin{itemize}
\item $\mathsf{Q_b' : b \to Set}$;
\item $\mathsf{L_E' : Equal^{\wedge}\,a\,(b \times b)\,Q_a'\,(Pair^{\wedge}\,b\,b\,Q'_b\,Q'_b)\,e}$;
\item $\mathsf{L_G' : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b')\,x}$;
\end{itemize}
%\begin{gather*}
%\mathsf{Q_a\;Q_a' : a \to Set} \\
%\mathsf{Q_b\;Q_b' : b \to Set} \\
%%\mathsf{M : PredMap\,b\,Q_a\,Q_a'} \\
%\mathsf{L_E : Equal^{\wedge}\,a\,(b \times b)\,Q_a\,(Pair^{\wedge}\,b\,b\,Q_b\,Q_b)\,e} \\
%\mathsf{L_G\ : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b)\,x} \\
%\mathsf{L_E' : Equal^{\wedge}\,a\,(b \times b)\,Q_a'\,(Pair^{\wedge}\,b\,b\,Q'_b\,Q'_b)\,e} \\
%\mathsf{L_G' : G^{\wedge}\,(G\,b)\,(G^{\wedge}\,b\,Q_b')\,x}
%\end{gather*}
In other words, we have a proof $\mathsf{L_E}$ of the (extensional) equality
of the predicates $\mathsf{Q_a}$ and $\mathsf{Pair^{\wedge}\,b\,b\,Q_b\,Q_b}$
and a morphism of predicates $\mathsf{M}$ from $\mathsf{Q_a}$ to $\mathsf{Q_a'}$,
and we need to use those to deduce a proof of the (extensional) equality
of the predicates $\mathsf{Q_a'}$ and $\mathsf{Pair^{\wedge}\,b\,b\,Q_b'\,Q_b'}$,
for some  for some predicate $\mathsf{Q_b'}$ on $\mathsf{b}$.
But that is not generally possible:
the facts that $\mathsf{Q_a}$ is equal to $\mathsf{Pair^{\wedge}\,b\,b\,Q_b\,Q_b}$
and that there is a morphism of predicates $\mathsf{M}$ from $\mathsf{Q_a}$ to $\mathsf{Q_a'}$
do not guarantee that $\mathsf{Q_a'}$ is equal to $\mathsf{Pair^{\wedge}\,bb\,Q_b'\,Q_b'}$ for some $\mathsf{Q_b'}$.

>>>>

  \begin{adjustbox}{varwidth=6.8in, max width=6.6in, margin=-0.1in 0in
      -0.2in 0in, fbox, center} 
{\small
\[\begin{array}{lll}
\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
  (var\;s\,T_A) \, liftA} & = & \mathsf{cvar \, A\, Q_A\, s\, T_A\,
  liftA}\\ 
\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
  (abs \,B \,C \,e \,s \,T_B \, t_C) \, (Q_B , Q_C , liftE,
  lift_{T_B}, lift_{t_C})} & = & \mathsf{cabs\,A\,B\,C\, Q_A\,
  Q_B\, Q_C\, e\, s\, T_B\, t_C\, liftE\, lift_{T_B}\, p_C}\\
\mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clist \, A\, Q_A\,
    (app \,B \,\,t_{BA} \, t_B)\, (Q_B , list_{t_{BA}}, list_{t_B})} &
= & \mathsf{capp\,A\,B\,Q_A\, Q_B\, t_{BA}\, t_B\, p_{BA} \, p_B}\\
  \mathsf{dIndLTerm \, P\, cvar \, cabs\, capp\, clistc \, A\, Q_A\,
    (list \,B \,e \, ts) \, (Q_B , liftE', lift_{List})} & = & 
  \mathsf{clistc \,A\,B\,Q_A\, Q_B\, e\, ts\, liftE'\, p_{List} }
\end{array}\]}

@@

We can generalize the approach taken in
Section~\ref{sec:deep-ind-GADTs} to a general framework for deriving
deep induction rules for deep GADTs. We will treat deep GADTs of the
form 
\begin{equation}\label{eq:gadts}
\begin{array}{l}
  \mathsf{data\ G : Set^\alpha
    \to Set\ where}\\
\mathsf{\;\;\;\;\;\;\;\;c\, :\, F\,G\,\ol{B} \to G (\ol{K\,\ol{B}})}
\end{array}
\end{equation}
For brevity and clarity we indicate only one constructor $\mathsf{c}$
in~\eqref{eq:gadts}, even though a GADT can, in general, have any
finite number of them, each with a type the same form as
$\mathsf{c}$'s. In~\eqref{eq:gadts}, $\mathsf{F}$ and $\mathsf{K}$ are
type constructors with signatures $\mathsf{(Set^{\alpha} \to Set) \to
  Set^{\beta} \to Set}$ and $\mathsf{Set^{\beta} \to Set}$,
respectively. If $\mathsf{T}$ has type signature $\mathsf{Set^{\gamma}
  \to Set}$ then we say that $\mathsf{T}$ is a {\em type constructor
  of arity $\mathsf{\gamma}$}.  The overline notation denotes a finite
list whose length is exactly the arity of the type constructor being
applied to it. The number of type constructors $\mathsf{K}$ must
therefore be $\alpha$. The type constructor $\mathsf{F}$ must be
constructed inductively according to the following grammar:
\[\mathsf{F\,G\,\ol{B} :=
F'\,G\,\ol{B} \times F''\,G\,\ol{B}
\ \vert\ F'\,G\,\ol{B} + F''\,G\,\ol{B}
\ \vert\ F'\,\ol{B} \to F''\,G\,\ol{B}
\ \vert\ G(\ol{F'\,\ol{B}})
\ \vert\ H\,\ol{B}
\ \vert\ H (\ol{F'\,G\,\ol{B}})
}\]

{\color{red} HERE!}  This grammar is subject to the following
restrictions. In the third clause the type constructor $\mathsf{F'}$
does not contain $\mathsf{G}$. In the fourth clause, none of the
$\mathsf{\alpha}$-many type constructors in $\mathsf{\ol{F'}}$
contains $\mathsf{G}$.  This prevents nesting, which would make it
impossible to give an induction rule; see
Section~\ref{sec:GADT-nested} below. In the fifth clause, $\mathsf{H}$
does not contain $\mathsf{G}$. This subsumes the case in which
$\mathsf{F}$ is a closed type
%, so, in particular, the unit and empty types, $\mathsf{1}$ and
%$\mathsf{0}$,
{\color{red} and the case in which $\mathsf{F}$ is a variable.}  In
the sixth clause, $\mathsf{H : Set^\gamma \to Set}$ does not contain
$\mathsf{G}$.  Moreover, although $\mathsf{H}$ can be

construct

any (truly)
nested type, it cannot

construct

be a GADT.  This ensures that $\mathsf{H}$
admits functorial semantics~\cite{jp20}, and thus that
$\mathsf{H^{\wedge}}$, admits a map function
\[\mathsf{HLMap : \forall (\ol{A : Set}) (\ol{Q\;Q' : A \to Set}) 
\to \ol{PredMap\,A\,Q\,Q'} \to
PredMap\,(H\,\ol{A})\,(H^{\wedge}\,\ol{A}\,\ol{Q})\,
(H^{\wedge}\,\ol{A}\,\ol{Q'})}\] A concrete way to define
$\mathsf{HLMap}$ is by induction on the structure of the type
$\mathsf{H}$.
%, and by giving an inductive definition when $\mathsf{H}$
%is an ADT or a nested type.
But we omit such details since they are not essential to the present
discussion.

A further requirement that applies to all of the types appearing
above, including the $\mathsf{K}$s, is that every type needs to have a
predicate lifting function.  This is not an overly restrictive
condition, though: all types made by sums, products, arrow types and
type application do, and so do GADTs as defined above. A concrete way
to define the predicate-lifting function for a type is to proceed by
induction on the structure of the type, and we have seen in the
previous sections examples of how to do so for products and type
application.  We do not give here the general definition of lifting,
as that would require to first present a full type calculus, and that
is beyond the scope of the paper.

For ease of notation we assume in the development below that
$\mathsf{G}$ is a unary type constructor, i.e., that $\alpha = 1$
in~\eqref{eq:gadts}. Extending the argument to GADTs of arbitrary
arity presents no difficulty other than heavier notation. The
constructor $\mathsf{c}$ there can be rewritten using the
$\mathsf{Equal}$ type as
\[\mathsf{c : \forall (\ol{B : Set}) \to Equal\,A\,(K\,\ol{B}) \to
  F\,G\,\ol{B} \to G\,A}\]
which is the form we shall use from now on.

@@

In the previous sections,
we derived induction rules for GADTs that do not feature nesting,
in the sense that their constructors contain no nested calls of the recursive variable,
as truly nested types (such as $\mathsf{Bush}$, Equation~\ref{eq:bush}) do,
or other GADTs calling the recursive variable.
Since both nested types
and GADTs without nesting admit induction rules,
it seems reasonable to expect that GADTs with nesting would as well.
Surprisingly, that is not the case:
indeed, the induction principle generally relies on (unary) parametricity of the semantic interpretation,
and in the case of nested types it also relies on functorial semantics~\cite{jp20},
but GADTs cannot admit both functorial and parametric semantics at the same time~\cite{jgj21}.
In this section we show how induction for GADTs featuring nesting goes wrong
by analyzing the following concrete example of such a type.
\begin{equation}\label{gadt-nested}
\begin{array}{l}
\mathsf{data\ G : Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;\;\;\;c :\, G\,(G\,A) \to G\,(A \times A)}
\end{array}
\end{equation}
The constructor $\mathsf{c}$ can be rewritten as
\[
\mathsf{c : \forall\, (B : Set) \to Equal\,A\,(B \times B) \to G\,(G\,B) \to G\,A}
\]
which is the form we shall use from now on.
The predicate-lifting function of $\mathsf{G}$,
\[
\mathsf{G^{\wedge} : \forall\, (A : Set) \to (A \to Set) \to G\,A \to Set}
\]
is defined as
\[
\mathsf{G^{\wedge}\,A\,Q\,(c\,B\,e\,x)
= \sum\, [Q' : B \to Set]\,
Equal^{\wedge}\,A\,(B \times B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e
\times G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q')\,x}
\]
where $\mathsf{Q : A \to Set}$, $\mathsf{e : Equal\,A\,(B \times B)}$ and $\mathsf{x : G(G\,B)}$.
Finally, let $\mathsf{dIndC}$ be the function
\[
\begin{array}{l}
\mathsf{\lambda\, (P : \forall\, (A : Set) \to (A \to Set) \to G\,A \to Set)} \\
\quad\mathsf{\to \forall\, (A\;B : Set)\, (Q : A \to Set)\, (Q' : B \to Set)\, (e : Equal\,A\,(B \times B))\, (x : G\,(G\,B))} \\
\quad\mathsf{\to Equal^{\wedge}\,A\,(B \times B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e
	\to P\,(G\,B)\,(P\,B\,Q')\,x
	\to P\,A\,Q\,(c\,B\,e\,x)}
\end{array}
\]
associated to the $\mathsf{c}$ constructor.

The induction rule for $\mathsf{G}$ is
\[
\mathsf{\forall\, (P : \forall\, (A : Set) \to (A \to Set) \to G\,A \to Set)
\to dIndC\,P
\to \forall\, (A : Set)\, (Q : A \to Set)\, (y : G\,a) \to G^{\wedge}\,A\,Q\,y \to P\,A\,Q\,y}
\]
Consistently with the previous examples,
to prove the induction rule
we try to define a term of the above type, $\mathsf{dIndG}$, as
\[
\mathsf{dIndG\,P\,cc\,A\,Q\,(c\,B\,e\,x)\,(Q', liftE, liftG)
= cc\,A\,B\,Q\,Q'\,e\,x\,liftE\,p}
\]
where $\mathsf{cc : dIndC\,P}$,
$\mathsf{Q : A \to Set}$ and
\begin{itemize}
\item $\mathsf{c\,B\,e\,x : G\,A}$,
i.e., $\mathsf{e : Equal\,A\,(B \times B)}$ and $\mathsf{x : G\,(G\,B)}$;
\item $\mathsf{(Q', liftE, liftG) : G^{\wedge}\,A\,Q\,(c\,B\,e\,x)}$,
i.e., $\mathsf{Q' : B \to Set}$, $\mathsf{liftE : Equal^{\wedge}\,A\,(B \times B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e}$,
and $\mathsf{liftG : G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q')\,x}$.
\end{itemize}
We still need to define $\mathsf{p : P\,(G\,B)\,(P\,B\,Q')\,x}$.
We do so by using the induction rule and letting
\[
\mathsf{p = dIndG\,P\,cc\,(G\,B)\,(P\,B\,Q')\,x\,q}
\]
where we still need to provide $\mathsf{q : G^{\wedge}\,(G\,B)\,(P\,B\,Q')\,x}$.
If we had the map function of $\mathsf{G^{\wedge}}$,
\[
\mathsf{GLMap : \forall\, (A : Set)\, (Q\;Q'' : A \to Set) \to PredMap\,A\,Q\,Q'' \to PredMap\,(G\,A)\,(G^{\wedge}\,A\,Q)\,(G^{\wedge}\,A\,Q'')}
\]
then we would be able to define
\[
\mathsf{q = GLMap\,(G\,B)\,(G^{\wedge}\,B\,Q')\,(P\,B\,Q')\,(dIndG\,P\,cc\,B\,Q')\,x\,liftG}
\]
Unfortunately, we cannot define such a $\mathsf{GLMap}$.
Indeed, its definition would have to be
\[
\mathsf{GLMap\,A\,Q\,Q''\,m\,(c\,B\,e\,x)\,(Q', liftE, liftG) = (Q''', liftE', liftG')}
\]
for some $\mathsf{(Q''', liftE', liftG') : G^{\wedge}\,A\,Q''\,(c\,B\,e\,x)}$, i.e.,
\begin{itemize}
\item $\mathsf{Q''' : B \to Set}$;
\item $\mathsf{liftE' : Equal^{\wedge}\,A\,(B \times B)\,Q''\,(Pair^{\wedge}\,B\,B\,Q'''\,Q''')\,e}$;
\item $\mathsf{liftG' : G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q''')\,x}$;
\end{itemize}
where $\mathsf{Q : A \to Set}$, $\mathsf{Q'' : A \to Set}$,
$\mathsf{m : PredMap\,A\,Q\,Q''}$,
$\mathsf{c\,B\,e\,x : G\,A}$, i.e.,
\begin{itemize}
\item $\mathsf{e : Equal\,A\,(B \times B)}$;
\item $\mathsf{x : G\,(G\,B)}$;
\end{itemize}
and $\mathsf{(Q', liftE, liftG) :G^{\wedge}\,A\,Q\,(c\,B\,e\,x)}$, i.e.,
\begin{itemize}
\item $\mathsf{Q' : B \to Set}$;
\item $\mathsf{liftE : Equal^{\wedge}\,A\,(B \times B)\,Q\,(Pair^{\wedge}\,B\,B\,Q'\,Q')\,e}$;
\item $\mathsf{liftG : G^{\wedge}\,(G\,B)\,(G^{\wedge}\,B\,Q')\,x}$.
\end{itemize}
In other words, we have a proof $\mathsf{liftE}$ of the (extensional) equality
of the predicates $\mathsf{Q}$ and $\mathsf{Pair^{\wedge}\,B\,B\,Q'\,Q'}$
and a morphism of predicates $\mathsf{m}$ from $\mathsf{Q}$ to $\mathsf{Q''}$,
and we need to use those data to deduce a proof of the (extensional) equality
of the predicates $\mathsf{Q''}$ and $\mathsf{Pair^{\wedge}\,B\,B\,Q'''\,Q'''}$,
for some predicate $\mathsf{Q'''}$ on $\mathsf{B}$.
But that is not generally possible:
the facts that $\mathsf{Q}$ is equal to $\mathsf{Pair^{\wedge}\,B\,B\,Q'\,Q'}$
and that there is a morphism of predicates $\mathsf{m}$ from $\mathsf{Q}$ to $\mathsf{Q''}$
do not guarantee that there exists a predicate $\mathsf{Q'''}$ such that $\mathsf{Q''}$ is equal to $\mathsf{Pair^{\wedge}\,B\,B\,Q'''\,Q'''}$.

At a deeper level, the fundamental issue is that
the $\mathsf{Equal}$ type does not have functorial semantics,
so that having morphisms $\mathsf{A \to A'}$ and $\mathsf{B \to B'}$
(for any type $\mathsf{A, A', B}$ and $\mathsf{B'}$)
and a proof that $\mathsf{A}$ is equal to $\mathsf{A'}$
does not provide a proof that $\mathsf{B}$ is equal to $\mathsf{B'}$.
This is because GADTs can either have a syntax-only semantics or a functorial-completion semantics.
Since we are interested in induction rules, we consider the syntax-only semantics,
which is parametric but not functorial.
Had we considered the functorial-completion semantics, which is functorial,
we would have forfeited parametricity instead.
In both cases, thus, we cannot derive an induction rule for GADTs featuring nesting.
Unlike nested types, indeed, GADTs do not admit a semantic interpretation that is both parametric and functorial~\cite{jgj21}.

\section{Applications}

%We want a proof
%\[
%  \mathsf{getTypeProof : \forall \, (A : Set) \to LTerm\,A \to Maybe \, (LType \, A)} \\
%\]
%that takes a lambda term and produces its type (using $\mathsf{Maybe}$ to represent
%potential failure).
%Because of the $\mathsf{listC}$ constructor of $\mathsf{LTerm}$, this cannot be achieved without 
%deep induction. In particular, deep induction is required to apply the induction to the 
%individual terms in a list of terms.

In this section we use deep induction for the $\mathsf{LTerm}$ GADT from Example~\ref{example:lterm} to 
extract the type from a lambda term.
Consider the predicate 
\begin{align*}
  &\mathsf{GetType : \forall \, (A : Set) \to LTerm\,A \to Set} \\
  &\mathsf{GetType \,A \,t = Maybe \, (LType \, A)}
\end{align*}
that takes a lambda term and produces the type of its possible types.
This predicate uses $\mathsf{Maybe}$ to represent
potential failure, where $\mathsf{Maybe}$ is the type defined as follows:
\begin{equation}\label{eq:maybe}
\mathsf{data\ Maybe : Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;nothing :\, Maybe\,A}\\
\mathsf{\;\;\;\;\;\;\;\;\;\;\;\;just :\, A \to Maybe\,A}
\end{equation}
We want to show this predicate is satisfied for every element of $\mathsf{LTerm\,A}$,
i.e., we want to prove: 
